---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    keep_md: yes
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(reshape2)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

Generalizability: 
Each survey from 1972 to 2004 was an independently drawn sample of English-speaking persons 18 years of age or over, living in non-institutional arrangements within the United States. Starting in 2006 Spanish-speakers were added to the target population. Full probability sampling (with random selection) was employed only in half of the 1975 and 1976 surveys and then 1977 onward. Before this, block quota sampling was used. 

Since the sample for the GSS is a cluster sample, standard errors are larger for the GSS than simple random sample calculations. Therefore, without applying the appropriate corrections for the design, there are reservations in generalizing to the entire population. 

Causality: 
Since this is simply an observational study, with no treatment/random assignment applied to the population, causality cannot be interpreted from the variables.  

* * *

## Part 2: Research question

> Is there a relationship between a person's opinion on abortion (ABANY) and   
a) Their highest level of education (DEGREE)  
b) Political party affiliation (PARTYID)  
c) How often someone attends religious service (ATTEND)  

The reason this research question is important to me is because of two reasons: 
1) I am strong advocate for women's rights and am pro-abortion
2) States across the U.S passed bills in 2019 make abortion past six weeks into pregnancy illegal

I am interested in researching whether anti-abortion opinion and education/party affiliation/religion are dependent. The reason that I have chosen these three variables to conduct my hypothesis tests is because I want to understand this demographic on three levels - personal (highest level of education), political (party), and social (religion). This will help with creating effective communication strategies (the targeting, the messaging) to inform people on this important women's right. 

* * *

## Part 3: Exploratory data analysis


```{r}
data <- gss %>% 
  select(year,abany,degree,partyid,relig,attend) %>%
  filter(!is.na(degree)) %>% 
  filter(!is.na(partyid)) %>%
  filter(!is.na(relig)) %>% 
  filter(!is.na(attend)) %>% 
  filter(!is.na(abany)) %>%
  mutate(antiabcode = if_else((abany=="No"),1,0))

degree_summary <- data %>% 
  select(degree,antiabcode) %>%
  group_by(degree) %>%
  summarize_at(vars(antiabcode),funs(prop_anti_abortion=mean,count=n()))

degree_summary

data %>% group_by(degree,abany) %>% 
  summarise(n=n()) %>% mutate(pct = n/sum(n)) %>%
  ggplot(aes(x=degree,y=pct,fill=abany,label=paste0(round(pct*100),"%"))) + 
  geom_col(position= position_stack(), color = "black") + 
  geom_text(position=position_stack(vjust=.5)) + 
  coord_flip() + 
  scale_y_continuous(labels=scales::percent_format()) + 
  xlab("Highest Level of Education") + 
  ylab("Proportion of People Against Abortion") + 
  labs(title="Relationship between Education and Opinion on Abortion") + 
  scale_fill_discrete(name = "Abortion", labels = c("Pro","Anti"))

```

We can see that the higher the educational degree achieved, the lesser the proportion of people who are anti-abortion. 

```{r}
party_summary <- data %>% 
  select(partyid,antiabcode) %>%
  group_by(partyid) %>%
  summarize_at(vars(antiabcode),funs(prop_anti_abortion=mean,count=n())) 
  

party_summary$partyid <- factor(party_summary$partyid,levels = party_summary$partyid[order(party_summary$prop_anti_abortion)])

party_summary


ggplot(party_summary,aes(x=partyid,y=prop_anti_abortion))+
  theme_bw()+ 
  geom_bar(stat="identity")+
  coord_flip()+
  xlab("Party Affiliation") + 
  ylab("Proportion of People Against Abortion") + 
  labs(title="Relationship between Party Affiliation and Opinion on Abortion")+
  geom_text(aes(label=paste0(round(prop_anti_abortion*100),"%")),hjust=+1.2,colour="white") +
  scale_y_continuous(labels = scales::percent)


```


Thus, we can see that democrats are on the lower end of the spectrum when it comes to being anti-abortion, whereas Republicans are at the higher end. Surprisingly, Independent party affiliation has the second highest proportion of people against abortion. 

```{r}
rel_summary <- data %>% 
  select(attend,antiabcode) %>%
  group_by(attend) %>%
  summarize_at(vars(antiabcode),funs(prop_anti_abortion=mean,count=n())) 
  

rel_summary$attend <- factor(rel_summary$attend,levels = rel_summary$attend[order(rel_summary$prop_anti_abortion)])

rel_summary


ggplot(rel_summary,aes(x=attend,y=prop_anti_abortion))+
  theme_bw()+ 
  geom_bar(stat="identity")+
  coord_flip()+
  xlab("Frequency of attending religious service") + 
  ylab("Proportion of People Against Abortion") + 
  labs(title="Relationship between Religiosity and Opinion on Abortion")+
  geom_text(aes(label=paste0(round(prop_anti_abortion*100),"%")),hjust=+1.2,colour="white") +
  scale_y_continuous(labels = scales::percent)
```


There is clear correlation (not causation) between religiosity and opinion on Abortion, where higher frequency of attending religious service is associated with a larger proportion of anti-abortion opinions. 

* * *

## Part 4: Inference

We will use the chi-square test for a two way table to test independence in each of the three cases. The chi-square test calculates a chi-square test statistic and a corresponding p-value. The p-value is the probability of obtaining a sample with observed counts given that the null hypothesis is true. If the p-value is lower than 0.05, we are able to reject the null hypothesis. 

Since I am not calculating the difference between 2 proportions, rather researching the dependency between two variables, I have not created confidence intervals. 

> 1) Is there a relationship between abortion opinion and highest level of education achieved? 

The hypotheses are as follows: 

H0: Abortion opinion and education are independent variables  
HA: Abortion opinion and education are dependent variables  

Conditions to check before performing chi-square test:  
a) Independence - Each case that contributes a count to the table must be independent of all other cases in the table   
b) Sample size - Each cell must have at least 5 expected cases  

We can confirm that the respondents have been randomly sampled, therefore each case is independent of the others. Sample size condition is satisfied as well, as seen from the table of counts below.  

```{r}
degree_table <- data %>% 
  select(degree,abany) 
  
table(degree_table$degree,degree_table$abany)

chisq.test(degree_table$degree,degree_table$abany,correct=FALSE)
```

Since p-value is lesser than 0.05, we can reject the null hypothesis and conclude that the data provides strong evidence that the highest level of education and opinion on abortion are dependent. In other words, given that the variables are independent, the probability of collecting a sample with the observed counts is almost 0. 


> 2) Is there a relationship between abortion opinion and party affiliation? 

The hypotheses are as follows: 

H0: Abortion opinion and party affiliation are independent variables  
HA: Abortion opinion and party affiliation are dependent variables   

Conditions to check before performing chi-square test:  
a) Independence - Each case that contributes a count to the table must be independent of all other cases in the table  
b) Sample size - Each cell must have at least 5 expected cases  

We can confirm that the respondents have been randomly sampled, therefore each case is independent of the others. Sample size condition is satisfied as well, as seen from the table of counts below.  

```{r}
party_table <- data %>% 
  select(partyid,abany) 
  
table(party_table$partyid,party_table$abany)

chisq.test(party_table$partyid,party_table$abany,correct=FALSE)
```


Since p-value is lesser than 0.05, we can reject the null hypothesis and conclude that the data provides strong evidence that party affiliation and opinion on abortion are dependent. In other words, given that the variables are independent, the probability of collecting a sample with the observed counts is almost 0. 


> 3) Is there a relationship between abortion opinion and frequency of attending religious service? 

The hypotheses are as follows: 

H0: Abortion opinion and frequency of attending religious service are independent variables  
HA: Abortion opinion and frequency of attending religious service are dependent variables   

Conditions to check before performing chi-square test:  
a) Independence - Each case that contributes a count to the table must be independent of all other cases in the table  
b) Sample size - Each cell must have at least 5 expected cases  

We can confirm that the respondents have been randomly sampled, therefore each case is independent of the others. The sample size condition is also met as seen from the table below.  

```{r}
relig_table <- data %>% 
  select(attend,abany) 

table(droplevels(relig_table$attend),relig_table$abany)

chisq.test(relig_table$attend,relig_table$abany,correct=FALSE)
```

Since p-value is lesser than 0.05, we can reject the null hypothesis and conclude that the data provides strong evidence that frequency of attending religious service and opinion on abortion are dependent. In other words, given that the variables are independent, the probability of collecting a sample with the observed counts is almost 0. 